Joe Biden Predicted to Win 55.9% of the Popular Vote in the 2024 U.S. Presidential Election

### 2024-03-16 paper/paper.qmd

---
title: "Joe Biden Predicted to Win 55.9% of the Popular Vote in the 2024 U.S. Presidential Election"
author: 
  - Talia Fabregas
  - Fatimah Yunusa
  - Aamishi Sundeep Avarsekar
thanks: "Code and data are available at: https://github.com/taliafabs/US-Election-Forecast-2024.git"
date: today
date-format: long
abstract: "The 2024 U.S. Presidential Election will take place on November 5th and it will be a rematch between President Joe Biden and Former President Donald Trump. In this report, we used a logistic regression model along with multi-level post-stratification (MRP) to predict the winner of the upcoming election. Using the results of our model, we predict that President Biden will win 56.59% of the popular vote and defeat former President Trump 467 to 71 in the electoral college, and win a second term in the White House. However, due to the limited nature of our survey data, the error range of our electoral college prediction is large."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Donald Trump refused to acknowledge his loss to Joe Biden in the 2020 U.S. presidential election. Since November 2020, he has falsely claimed that the election was fraudulent, and that Biden stole the election from him. On January 6, 2021, an angry pro-Trump mob stormed the U.S. Capitol in an attempt to "stop the steal," or in other words, to stop the certification of Biden's electoral college victory and overturn the results of the election.

@@

Clear gap that needs to be filled ... what is the research gap and why is this important? - 2024 US Presidential Election - how consequential this election is - what is on the ballot - women's rights - trump's project 2025 thing.

If someone's done it before not on this data set not in this context.

@@

Our report has 5 sections. In @sec-data, we examine the data sets that we used for our report. In @sec-model, we we use multi-level regression with post-stratification (MRP) to predict the results of the 2024 U.S. presidential election. MRP uses two different data sets, our survey data and our post-stratification data. It fits a model using the smaller survey data set and applies it to the larger post-stratification data set. In @sec-results, we present the findings of our model. In @sec-discussion, we discuss what we learned.

@@

In @fig-survey_presidential_preferences1, we can see that support for Biden within our survey data varies by race and sex. Black women showed overwhelming support for President Biden's upcoming re-election campaign, while white men appear to be split between Biden and Trump. As shown in @tbl-racetables 69% of respondents were white. Among white survey respondents, women were more likely to support President Biden and men were more likely to support Trump. We used code from Chapter 13 of Telling Stories with Data and modified it to produce the bar graphs in this section [@tswd].

@@

Using only the survey data, we can calculate the overall support for Biden and the support for Biden in each state to make a popular vote and electoral college prediction. Based on our earlier findings, we believe that being non-white, female, college or post-grad educated, and living in an urban area makes an individual more likely to support Biden. To calculate a popular vote prediction for President Biden based on the survey data, we simply calculated the mean of the `vote_biden` binary indicator variable, and to calculate an electoral college prediction, we added up the electoral college votes of the states where Biden received more than 50% support from survey respondents. We used the steps and code provided by @citeMatthew and the `statebins` package [@statebins] to produce @fig-electoral-map-survey.

@@

@@

-   `hispan`: whether the respondent is of Hispanic origin or not.

@@

The sex, race, educ, and metro variables were cleaned and re-factored to mirror our survey data and to allow us to apply our model to our post-stratification data. @fig-demographics1 shows the demographic statistics of our survey and post-stratification data. The steps and code for producing @fig-demographics1 were obtained from @citeMatthew. We also created a `race_hispanic`, which is equal to either "hispanic" or "not hispanic" to ensure that our survey and post-stratification data sets would align, because this was a separate variable from race in our post-stratification data.

@@

demographic_variables |> ggplot(aes(as.factor(level), pct, group=as.factor(type), linetype = as.factor(type))) +
  geom_line() + facet_grid(~group, scales = "free") +
  theme(axis.text.x = element_text(angle=70, size = 6, hjust = 1), legend.position = "bottom") +
  labs(x = "Category", y = "Proportion", linetype = "data set") +
  scale_y_continuous(labels = scales::percent) +
  geom_hline(yintercept = 0.5, alpha = 0.35) 
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: true
#| label: tbl-racetables
#| tbl-cap: Races of survey and post stratification data respondents
#| tbl-subcap: ["Survey data", "Post-stratification data"]
#| layout-ncol: 2

races_table <- survey_analysis_data |>
  group_by(races) |>
   summarise(Count = n()) |>
  mutate("%" = round(Count / nrow(survey_analysis_data) * 100, 2))

kable(races_table)

races2 <- poststrat_analysis_data |>
  group_by(races) |>
  summarise(Count = n())|>
  mutate("%" = round(Count / nrow(poststrat_analysis_data) * 100, 2))

races2 <- races2 |> rename(Race = races)

@@

Overall, race demographics are consistent between our survey and post-stratification data sets. @tbl-racetables shows the exact count and proportion of respondents of different races in each data set. The proportion of Hispanic respondents in our survey and post-stratification data is comparable, at 12.7% and 13.5%, respectively. While there is a higher proportion of Hispanic respondents in our post-stratification data set, the difference is less than one percentage point. The proportions of white respondents in our survey (69.4%) and post-stratification (69.5%) data sets are nearly identical. Black respondents make up

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| label: fig-stateprops
#| fig-cap: Survey and Post-Stratification Data Proportion of Voters by State

@@

We can see that there are some slight differences in the proportions of respondents in the survey and post-stratification data from Alabama, California, Florida, Kansas, Maryland, Massachussets, Michigan, New York, Nebraska, and Virginia.

@@

# Model {#sec-model}
We will use multilevel regression with post-stratification (MRP) for our analysis.

## Model set-up
To predict whether a person will vote for Joe Biden or Donald Trump, we fitted a logistic regression model using data from the America's Political Pulse Survey [@americaspoliticalpulse] and post-stratified it using ACS census data [@citeIPUMS]. Logistic regression only works for binary classification, therefore we created a `vote_biden` variable during our data cleaning process, which is equal to 1 if a person's preferred candidate is Joe Biden, and 0 if it is Donald Trump.

We are interested in estimating We fitted a logistic regression model used to predict support for Biden using sex, age, race, race_hispanic, highest level of education, state, and urban as predictors on the America's Political Pulse survey data. We then applied it to the ACS 2022 census data. Our model is as follows:

@@

# Results {#sec-results}

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| label: tbl-popvotepost-popularvote
#| tbl-cap: "2024 U.S. election result estimates for Joe Biden based on post-stratification"
#| tbl-subcap: ["Popular Vote", "Electoral College"]

@@

# electoral college
# had chatgpt make me an electoral college csv
electoral_college <- read_csv("electoral_college_votes.csv", show_col_types = FALSE)

@@

biden_ec_poststrat <- tibble(
  "Electoral College Estimate:" = c("Lower Estimate", "Mean Estimate", "Upper Estimate"),
  "Biden" = c(biden_ec_lower, biden_ec_mean, biden_ec_upper),
  "Trump" = c(538-biden_ec_lower, 538-biden_ec_mean, 538-biden_ec_upper)
)

@@

# popular vote table
kable(biden_popvote, align = "c") %>%
  kable_styling(bootstrap_options = "striped", full_width = F)

# electoral college table
kable(biden_ec_poststrat, align = "c") %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

@@

# Discussion {#sec-discussion}

@@

## Electoral College Projection {#sec-second-point}

## Large Variation {#sec-third-point}

@@

## Model Details

```{r}
#| echo: false
#| eval: true
#| warning: false

modelsummary::modelsummary(presidential_vote_model)
# looking at coefficient estimates using modelsummary()
```

@@

@fig-stanareyouokay-1 is a trace plot. It shows that the lines have a wide overlap with each other. This suggests that there is no poor convergence and the distribution behaves as expected.

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

>>>>

---
title: "Joe Biden Predicted to Win 55.9% of the Popular Vote in the 2024 U.S. Presidential Election"
author: 
  - Talia Fabregas
  - Fatimah Yunusa
  - Aamishi Sundeep Avarsekar
thanks: "Code and data are available at: https://github.com/taliafabs/US-Election-Forecast-2024.git"
date: today
date-format: long
abstract: "The 2024 U.S. Presidential Election will take place on November 5th and it will be a rematch between President Joe Biden and Former President Donald Trump. In this report, we used a logistic regression model along with multi-level post-stratification (MRP) to predict the winner of the upcoming election. Using the results of our model, we predict that President Biden will win 55.9% of the popular vote and defeat former President Trump 467 to 71 in the electoral college, and win a second term in the White House. However, due to the limited nature of our survey data, the error range of our electoral college prediction is large."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

@@

Talk about polarization in a discussion point instead of introduction

Clear gap that needs to be filled ... what is the research gap and why is this important? - 2024 US Presidential Election - how consequential this election is - what is on the ballot - women's rights - trump's project 2025 thing

If someone's done it before not on this data set not in this context

Previous groups have looked at ...

@@

Our report has 5 sections.

We are using multi-level regression with post-stratification (MRP) to predict the results of the 2024 U.S. presidential election. MRP uses two different data sets, survey data and post-stratification data. It fits a model using the smaller survey data set and applies it to the larger post-stratification data set.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| label: tbl-racessurvey
#| tbl-cap: Races of survey respondents

races_table <- survey_analysis_data |>
  group_by(races) |>
  summarize("%" = round(n() / nrow(survey_analysis_data) * 100, 2))

kable(races_table)
```

In @fig-survey_presidential_preferences1, we can see that support for Biden within our survey data varies by race and sex. Black women showed overwhelming support for President Biden's upcoming re-election campaign, while white men appear to be split between Biden and Trump. As shown in @tbl-racessurvey 69% of respondents were white. Among white survey respondents, women were more likely to support President Biden and men were more likely to support Trump. We used some of the code from Chapter 13 of Telling Stories with Data and modified it to produce the bar graphs in this section [@tswd].

@@

Using only the survey data, we can calculate the overall support for Biden and the support for Biden in each state to make a popular vote and electoral college prediction. Based on our earlier findings, we believe that being non-white, female, college or post-grad educated, and living in an urban area makes an individual more likely to support Biden. To calculate a popular vote prediction for President Biden based on the survey data, we simply calculated the mean of the `vote_biden` binary indicator variable, and to calculate an electoral college prediction, we added up the electoral college votes of the states where Biden received more than 50% support from survey respondents. We used the steps and code provided by @citeMatthew and the `statebins` package [@statebins] to produce @fig-electoral-map-survey .

@@

We selected variables that mirror the ones in our survey data set.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-racetable2
#| tbl-cap: Races of post stratification data

races2 <- poststrat_analysis_data %>% 
  group_by(races) %>% 
  summarise(n = n()) %>% 
  mutate(pct = round((n/sum(n)*100),2), type = "survey",
         group = "races")

races2 |> rename("Count" = n)

races2 |> rename(Race = races)

races2 |> rename("%" = pct)

@@

@@

The sex, race, educ, and metro variables were cleaned and re-factored to mirror our survey data and to allow us to apply our model to our post-stratification data. @fig-demographics1 shows the demographic statistics of our survey and post-stratification data. The steps and code for producing @fig-demographics1 were obtained from @citeMatthew.

@@

demographic_variables |> ggplot(aes(as.factor(level), pct, group=as.factor(type), linetype = as.factor(type))) +
  geom_line() + facet_grid(~group, scales = "free") +
  theme(axis.text.x = element_text(angle=70, size = 6, hjust = 1), legend.position = "bottom") +
  labs(x = "Category", y = "Proportion", linetype = "data set") +
  scale_y_continuous(labels = scales::percent) +
  geom_hline(yintercept = 0.5, alpha = 0.35)

@@

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-stateprops
#| tbl-cap: Survey and Post-Stratification Data Proportion of Voters by State

@@

However, we can see that there are some slight discrepancies in the proportions of respondents in the survey and post-stratification data from Alabama, California, Florida, Kansas, Maryland, Massachussets, Michigan, New York, Nebraska, and Virginia.

@@

# Model

We will use multilevel regression with post-stratification (MRP) for our analysis.

## Model set-up

To predict whether a person will vote for Joe Biden or Donald Trump, we fitted a logistic regression model using data from the America's Political Pulse Survey [@americaspoliticalpulse] and post-stratified it using ACS census data [@citeIPUMS]. Logistic regression only works for binary classification, therefore we created a `vote_biden` variable during our data cleaning process, which is equal to 1 if a person's preferred candidate is Joe Biden, and 0 if it is Donald Trump.

The logistic regression model used to predict support for Biden using sex, age, race, highest level of education, state, and whether the individual lives in an urban or rural area is as follows:

<small>

$$
Pr(vote\_biden=1) = \text{logit}^{-1}(\beta_0 + \beta_1 \text{sex}_i + \beta_2  \text{age\_bracket}_i + \beta_3 \text{race}_i + \beta_4 race_hispanic_i + \beta_5 \text{education\_level}_i + \beta_6 \text{urban}_i )
$$

</small>

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelsummary
#| tbl-cap: "Coefficient estimates"
#| warning: false

# looking at coefficient estimates using modelsummary()
modelsummary::modelsummary(presidential_vote_model)
```

# Results

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| label: tbl-popvotepost
#| tbl-cap: "Popular Vote Estimates for Joe Biden"

@@

# Render the table using kable
kable(biden_popvote, align = "c") %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

```{r}
#| eval: true
#| echo: false
#| warning: false
#|abel: tbl-ecpost
#| tbl-cap: "Estimates for the Electoral College based on Post-Stratification"

# had chatgpt make me an electoral college csv
electoral_college <- read_csv("electoral_college_votes.csv", show_col_types = FALSE)

@@

biden_ec_poststrat <- tibble(
  "Electoral College Estimate:" = c("Lower", "Mean", "Upper"),
  "Biden" = c(biden_ec_lower, biden_ec_mean, biden_ec_upper),
  "Trump" = c(538-biden_ec_lower, 538-biden_ec_mean, 538-biden_ec_upper)
)

@@

# Render the table using kable
kable(biden_ec_poststrat, align = "c") %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

@@

# Discussion

@@

## Electoral College Projection

## Large Variation

@@

# Model details {#sec-model-details}

type something here

@@

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", 
#| "Rhat plot"]
#| layout-ncol: 2
